Writing Trash Bin

### 2024-03-11 paper.qmd
added intro to vis bit and some changes to inference bit

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The lack of an econompasing definition is touched on by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement. In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertaitny visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.

# The Definition Problem
## Colloquial Definitions
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

@@

## Taxonomy Definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiatet. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

## Inference and Uncertainty
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

@@

# The Uncertainty Visualisation Problem

@@

## Definitions of "Uncertainty Visualisation" 
### Taxonomies of Uncertainty Visualisations
There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

@@

The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

### Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

@@

### Types of Uncertainty Visualisation Papers

@@

# The Measureing Noise Problem

## Uncertainty vs Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

Most research into general uncertainty communication seems to focus specifically on how we communicate risks. Risk and uncertainty are not perfect synonyms of one another. @Spiegelhalter2017 makes the distinction clear by specifying *risk* as *a precise random event* (such as a coin flip) while *uncertainty* (also sometimes referred to as abiguity) is imprecision (such as the estimated probability associated with a coin that might be biased). For this reason, risk is often seen as a specific case of uncertainty and that well defined sub area has a lot of detailled research. Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. The problem with a risk focus is that it is only a small piece of the uncertainty pie, and very few of these findings can be reasonably assumed to go beyond risk communication. Even less can be assumed to translate to uncertainty visualisation.

@@

>>>>

This review believe all these sources of noise have a single root cause, a lack of a cohesive and encompassing definition of uncertainty. The absence of an econompasing definition of uncertainty is mentioned by every uncertainty visualisation directly [@Spiegelhalter2017; @Griethe2006] or indirectly by describing a miriad of ways it can be considered in the literature [@Kinkeldey2014; @Hullman2016], although it is never commented on as a source of the noise in the field.

In understanding the definition problem, we will understand the *purpose* of visualising uncertainty, and understand if the current visualisation tools are sufficient to achieve that purpose and if not, where there is need for improvement. In understanding this purpose, we will also consider whether or not "uncertainty visualisation" is a valid sub-field of visualisation itself, or if that distinction has only arisen as a by product of the confused definition of uncertainty. Ultimately, this paper should serve as a guide for future uncertainty visualisation authors to prevent their work from becoming noise in what is already an incredibly noisy field.

# What is Uncertainty?
## Colloquial Definitions
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

@@

## Taxonomy Definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

## Defining uncertainty with respect to inference
What exactly is uncertainty, then? If we were to consider making this overarching definition, what would it need in order to be "encapsulating"? Well, let us consider some comments about uncertainty or common sentiments that come up time and time again in the literature.

We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics would we?

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field.
For example, do descriptive statistics have uncertainty? Descriptive statistics simply describe features of your sample as they are. The sample mean is the sample mean, the sample size is the sample size, and the sample variance is its sample variance. We would not say we are "uncertain" about these values unless we are questioning our own computation skills.

Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

Variance and probability are typically considered the closest of these concepts to uncertainty, however they clearly do not fully encapsulate the concept. @Spiegelhalter2017 makes the reason for this distinction clear, as in the absence of something to draw inference on, variance and probability describe precise random events (such as a coin flip), while uncertainty is imprecision (such as the estimated probability associated with a coin that might be biased). A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty.

@@

# What is an Uncertainty Visualisation?
The lack of a consistent definition of uncertainty means that every statement that builds upon "uncertainty" will be ill-formed. This includes statements about uncertainty visualisation. This appears in the literature as a mountain of seemingly conflicting statements. For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in most "uncertainty visualisation" experiments [] (*Cite Gap __*).

Attempts to rectify these conflicts always flutter about the surface and do not seem to grasp the core of this problem. For example, @wu2023rational noticed a series of papers with poorly defined tasks and suggested comparing study participants to a "rational agent" to help authors understand and remedy this issue. @Kim2019 believed an understanding of participants prior beliefs could assist in interpreting uncertainty visualisation results that tend to be incredibly noisy. What these suggestions fail to understand is that the entire practice of identifying "true" facts about "unertainty visualisation" is futile. It is impossible to make a statement about the properties of an "uncertainty visualisation" in the same way it is impossible to comment on $\frac 0 0 = 1$. Division by zero is undefined, so the statement is ill formed, similarly, "uncertainty" is not defined, so statements *about* "uncertainty" including statements about "uncertainty visualisation" are ill formed. Ill-formed statements can neither be true nor false and cannot be part of a logical argument, so experiments and research that build upon this ill-formed concept are neither true nor false, they simply exist. Someone who is less familiar with the rules of logical argument may beleive this understanding of the problem to an insufferable nitpick that creates impossible to live up to standards, however the consequences of this definitional issue are *very* real and costly.  The studies that investigate "uncertainty visualisation" may be identifying *something*, however there is good reason to believe that *something* is neither interesting nor does can it contribute to a larger theory of visualisation. This is something we illustrate in the following this section.

## Common misconceptions about uncertainty visualisation

### Defining "uncertainty visualisation" as one word

@@

There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

@@

### Misunderstanding the role of contextual information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

@@

@@

# Noise and Signal
## Uncertainty versus Risk
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

@@

## Suggestions for measuring uncertainty
### Types of Uncertainty Visualisation Papers

### 2024-03-11 writingdumpster.qmd
added intro to vis bit and some changes to inference bit

@@

# Main section outtakes
- tasks e.g. "What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g. "What is the noise around the estimate of the value of X"
Visualisation experiments are highly sensitive to the way questions are asked. For example, a study by @Hullman2015 that compared HOPs, error bar, and violin plots found that many participants reported that group A was more likely to be higher than group B, which was not not plausible given that the mean of B was larger than the mean of A. The authors attributed this obvious mistake to the visual representation, but we believe it was more likely caused by the distributions being shown in the order A then B while the question swapped the order by asking "What is P(B>A)?". This flip is minor, but must be done *after* the participants have worked out the probability. Since the primary task is done they forget to complete the secondary task, similar to forgetting your bank card in an ATM. I suspect this was the cause of the mistake as I, when attempting this question myself, made the exact same mistake. I also constalty lead my bank card in ATMs.

>>>>

- Some markers of probability that have common uses, probabilities such as 0 (impossible), 0.5 (a fair coin flip), 1 (certain) are easy for people to have an internal sense of *hullman paper*. 
- It is incredibly hard to communicate small risks, because people cannot differentiate between 1/100 an 1/1000, and communicating low probability but high impact risks are particularly challenging [@Spiegelhalter2017]. This issue can be somewhat alleviated by communicating the likelihood of an event in relation to another event on a similar scale (such as being struck by lightning) [@Spiegelhalter2017]. 
- The framing of how we communicate risks matters due to the affect heuristic, so there will be a difference in peoples behaviours if, for example, you tell people they have a 90% chance of survival vs a 10% chance of death, and when expressing a frequency, a larger numerator communicates a larger risk [@Spiegelhalter2017]. 
- The reference class of a particular risk should be explicitly stated to avoid confusion, for example if we are communicating the chance of rain, the time period over which we are expecting this chance of rain (i.e. 50% chance today or this week) needs to be specified [@Spiegelhalter2017]. 

@@

# Main section outtakes
- What exactly is uncertainty, then? Let us consider the concepts related to uncertainty that have been discussed so far. We know uncertainty cannot be entirely explained by individual concepts that *have* been defined, such as probability, variance, error, missing values, precision, or related concepts that pop up throughout the data analysis pipeline. We also know it likely has some relationship with the *task* we are considering, given that many authors have attributed the noise in uncertainty visualisation research as a type of "task" dependence. Additionally, it should be clear that we wouldn't consider uncertainty to be an appropriate term for *everything* in statistics. For example, I wouldn't describe the population parameters of a distribution I have simulated to be "uncertain"

- tasks e.g. "What is the average of X" (how well lined up you are moving down the data pipeline) and resolution (uncertainty is very often moving back a resolution level in statisitcs - also can be an assumption - bootstrap is a replacement for an assumption, not a resolution change, but bootstrapping is centered on the data)  e.g. "What is the noise around the estimate of the value of X"
Visualisation experiments are highly sensitive to the way questions are asked. For example, a study by @Hullman2015 that compared HOPs, error bar, and violin plots found that many participants reported that group A was more likely to be higher than group B, which was not not plausible given that the mean of B was larger than the mean of A. The authors attributed this obvious mistake to the visual representation, but we believe it was more likely caused by the distributions being shown in the order A then B while the question swapped the order by asking "What is P(B>A)?". This flip is minor, but must be done *after* the participants have worked out the probability. Since the primary task is done they forget to complete the secondary task, similar to forgetting your bank card in an ATM. I suspect this was the cause of the mistake as I, when attempting this question myself, made the exact same mistake. I also constalty lead my bank card in ATMs.
